---
title: "Humanitarian data analysis via humanitarian-mcp"
output: html_document
---

Four reproducible research workflows against a local `humanitarian-mcp`
server. Start it first: `npm run dashboard` (see `README.md` in this folder).

```{r setup}
library(httr2)
library(jsonlite)

SERVER <- "http://localhost:8642"

# One MCP tool call via the dashboard bridge; returns structuredContent.
call_tool <- function(tool, ...) {
  payload <- request(paste0(SERVER, "/api/call")) |>
    req_body_json(list(tool = tool, arguments = list(...))) |>
    req_perform() |>
    resp_body_json()
  if (isTRUE(payload$isError)) stop(payload$content[[1]]$text)
  payload$structuredContent
}

# export_data -> data.frame. The manifest rides in '#' comment lines.
export_csv <- function(...) {
  result <- call_tool("export_data", format = "csv", ...)
  read.csv(text = result$data, comment.char = "#")
}
```

## Workflow 1 — Event study: displacement from Sudan around April 2023

```{r event-study}
sudan <- export_csv(dataset = "population", country = "Sudan",
                    role = "origin", year_from = 2010)
yearly <- aggregate(refugees ~ year, data = sudan, FUN = sum)
yearly$post <- as.integer(yearly$year >= 2023)
yearly$t <- yearly$year - min(yearly$year)

summary(lm(refugees ~ t * post, data = yearly))
```

## Workflow 2 — Recognition rates across host states

```{r recognition-rates}
rates <- do.call(rbind, lapply(c("Egypt", "Germany", "Kenya"), function(country) {
  yearly <- call_tool("asylum_decisions", country = country, year_from = 2019)$yearly
  data.frame(
    country = country,
    year = sapply(yearly, `[[`, "year"),
    recognition_rate_pct = sapply(yearly, `[[`, "recognition_rate_pct")
  )
}))
reshape(rates, idvar = "year", timevar = "country", direction = "wide")
```

## Workflow 3 — Hosting burden per capita

```{r per-capita}
ranking <- call_tool("top_host_countries", normalize_by = "population", limit = 10)
per_capita <- data.frame(
  country = sapply(ranking$ranking, `[[`, "country"),
  value = sapply(ranking$ranking, `[[`, "value")
)
barplot(rev(per_capita$value), names.arg = rev(per_capita$country), horiz = TRUE,
        las = 1, xlab = paste("refugees", ranking$unit, sprintf("(%d)", ranking$year)))
```

## Workflow 4 — Conflict × displacement (needs the `hdx` provider)

```{r conflict, error=TRUE}
conflict_records <- call_tool("conflict_events", country = "Sudan", year_from = 2018)$records
conflict <- data.frame(
  year = sapply(conflict_records, `[[`, "year"),
  fatalities = sapply(conflict_records, `[[`, "fatalities")
)
displaced <- export_csv(dataset = "population", country = "Sudan",
                        role = "origin", year_from = 2018)
displaced_yearly <- aggregate(refugees ~ year, data = displaced, FUN = sum)
merged <- merge(conflict, displaced_yearly, by = "year")
merged
cor(merged$fatalities, merged$refugees)
```

## Method notes

- Figures are end-year stocks; UNHCR revises series retroactively — record your
  extraction date (the manifest in every export does this for you).
- `role = "asylum"` = hosted *in* the country; `role = "origin"` = displaced *from* it.
- Add `include_codebook = TRUE` to any `export_data` call for a variable-level codebook.
